---
title: "Cyclistic Case Study"
author: "Cameron"
date: '2023-03-01'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Cyclistic Capstone Project
This project is the capstone project of the [Google Data Analytics Certificate course](https://grow.google/certificates/data-analytics/#?modal_active=none). The scenario involves a fictional company, Cyclistic, a bike sharing service based in Chicago. The director of marketing believes the company’s future success depends on maximizing the number of annual memberships. Therefore,the team wants to understand how casual riders and annual members use Cyclistic bikes differently. From these insights,the team will design a new marketing strategy to convert casual riders into annual members. 

## Ask
The marketing team has come up with three questions that will need to be answered for this campaign. 
1. How do annual members and casual riders use Cyclistic bikes differently?
2. Why would casual riders buy Cyclistic annual memberships?
3. How can Cyclistic use digital media to influence casual riders to become members?
This report will be tackling the first question. 

## Prepare
The data used for this project comes from an actual company Divvy based in Chicago so it is primary data. They have done their own pre-cleaning to remove PII and trips less than 60 seconds but cleaning will still be necessary to ensure we have the best data and accurate analysis. As I am the only one involved in this project I will be responsible for the naming and storage of all the data.

## Process
The first thing I did was load the libraries necessary for cleaning, analysis, and visualizations

```{r libraries}
#Preparing the environment
library("ggplot2")
library("tidyverse")
library("ggmap")
library("ggthemes")
library("maps")
```

I then loaded all of the cvs files and named them to create tables
```{r loading cvs files}
#Bringing in the data        
feb_2022 <- read_csv("R/Projects/Cyclistic/202202-divvy-tripdata.csv")
mar_2022 <- read_csv("R/Projects/Cyclistic/202203-divvy-tripdata.csv")
apr_2022 <- read_csv("R/Projects/Cyclistic/202204-divvy-tripdata.csv")
may_2022 <- read_csv("R/Projects/Cyclistic/202205-divvy-tripdata.csv")
jun_2022 <- read_csv("R/Projects/Cyclistic/202206-divvy-tripdata.csv")
jul_2022 <- read_csv("R/Projects/Cyclistic/202207-divvy-tripdata.csv")
aug_2022 <- read_csv("R/Projects/Cyclistic/202208-divvy-tripdata.csv")
sep_2022 <- read_csv("R/Projects/Cyclistic/202209-divvy-publictripdata.csv")
oct_2022 <- read_csv("R/Projects/Cyclistic/202210-divvy-tripdata.csv")
nov_2022 <- read_csv("R/Projects/Cyclistic/202211-divvy-tripdata.csv")
dec_2022 <- read_csv("R/Projects/Cyclistic/202212-divvy-tripdata.csv")
jan_2023 <- read_csv("R/Projects/Cyclistic/202301-divvy-tripdata.csv")

```

Next I dropped the NULL rows from all of the tables.
```{r Dropping}
apr_2022 <- apr_2022 %>% drop_na()
aug_2022 <- aug_2022 %>% drop_na()
dec_2022 <- dec_2022 %>% drop_na()
feb_2022 <- feb_2022 %>% drop_na()
jan_2023 <- jan_2023 %>% drop_na()
jul_2022 <- jul_2022 %>% drop_na()
jun_2022 <- jun_2022 %>% drop_na()
mar_2022 <- mar_2022 %>% drop_na()
may_2022 <- may_2022 %>% drop_na()
nov_2022 <- nov_2022 %>% drop_na()
oct_2022 <- oct_2022 %>% drop_na()
sep_2022 <- sep_2022 %>% drop_na()
```

Then I brought all of the 12 months together under `cyc`

```{r together}
#Bringing it all together
cyc <- rbind(feb_2022, mar_2022, apr_2022, may_2022, jun_2022, jul_2022, aug_2022, sep_2022, oct_2022, nov_2022, dec_2022, jan_2023)
```

After that I doubled checked to make sure all null rows were gone, removed any duplicate rows/
```{r doublecheck}
#Checking if there are missing values
anyNA(cyc)
head (cyc)

#Removing any duplicate rows
cyc <- cyc[!duplicated(cyc$ride_id), ]
```

The last bit of processing before I got to the analysis was to add a column for the day of the week the ride started on and a column for ride length. I then cleaned all rows which had a ride length less than one minute.
```{r newcolumns}
#Adding a ride length column and a day of week column
cyclistic_clean <- cyc %>% mutate(ride_length = difftime(ended_at, started_at, units = "mins" ) ) %>% 
  mutate(day_of_week = weekdays(started_at))
head(cyclistic_clean)

#Last step of cleaning filtering out ride lengths less than one
cyclistic_clean <- cyclistic_clean %>% filter(ride_length >= 1 )

summary(cyclistic_clean)
```

## Analyze

Now I analyzed the data. I won't go into much detail because each line of code is explained. Some data visualizations are also included in the code.

```{r analysis, echo=FALSE}
#Changing the levels for the week data to appear in the correct order
cyclistic_clean$day_of_week <- factor(cyclistic_clean$day_of_week, levels= c("Sunday", "Monday", 
                                                                             "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
#Getting numbers of rides and a data visualization
cyclistic_clean %>% group_by(member_casual) %>% count(member_casual)

ggplot(data = cyclistic_clean) + geom_bar(mapping = aes(member_casual, fill = member_casual)) + 
  labs(title = "Number of rides in the past 12 months", x = "Member or Casual Rider", fill = "Key")

#Prepare data to find avg ride length
cyclistic_clean$ride_length <- as.numeric(cyclistic_clean$ride_length)

#Average ride length by day of week separated by member and casual
cyclistic_clean %>% group_by(member_casual, day_of_week) %>% 
  summarise(mean_ride_length=mean(ride_length),
            .groups = 'drop')

#Getting rides broken out by month
cyclistic_clean %>% 
  group_by(month = lubridate::floor_date(started_at, 'month'),member_casual) %>%
  count()

#Adding a month column to the data table
cyclistic_clean$month <- as.Date(cut(cyclistic_clean$started_at, breaks = 'month'))

#Making a table for the data visualization
avg_ride_month <- cyclistic_clean %>% group_by(member_casual, month) %>% 
  summarise(mean_ride_length=mean(ride_length),
            .groups = 'drop')

#Data visualization of mean ride length each month separated by member and casual
ggplot(avg_ride_month, aes(x=month, y=mean_ride_length)) + 
  geom_bar(aes(fill = member_casual),stat="identity", position = "dodge") + 
  labs(title = "Average ride length by month", fill = "Key", x = "Month", y = "Average ride length")

#Bar graph of rides each month broken out by members and casual riders
ggplot(cyclistic_clean, aes(x=month, after_stat(count))) + 
  geom_bar(aes(fill = member_casual), position = "dodge") + 
  labs(title = "Number of rides separated by month", fill = "Key", x = "Month")


#Counting the number or rides by weekday divided by member and casual
cyclistic_clean %>% group_by(member_casual) %>% count(day_of_week)


#Data Visualization for types of riders by day of the week
ggplot(cyclistic_clean, aes(x=day_of_week,after_stat(count) )) + 
  geom_bar(aes(fill = member_casual), position = "dodge") +
  labs(title = "Number of rides separated by day of week", x = "Day of week", fill = "Key")
  

#Which stations are most popular with casual riders and members?
cyclistic_clean %>% count(member_casual, start_station_name, sort = TRUE)
cyclistic_clean %>% count(member_casual, end_station_name, sort = TRUE)

#Preparing data for a map of the most popular stations
top_stations <-cyclistic_clean %>% count(member_casual, start_station_name, sort = TRUE) %>% slice_head(n=25)

top_stations_map <- top_stations %>% left_join(cyclistic_clean, by = join_by(start_station_name), multiple = "any") %>% 
  select(member_casual.x, start_station_name, n, start_lat, start_lng)

#Get the map for the data visualization
chicago_map <- get_stamenmap(
  bbox = c(left = -87.7, bottom = 41.7877 , right = -87.594, top = 41.9437), 
  maptype = "terrain",
  zoom = 11)
ggmap(chicago_map) + # creates the map "background"
  geom_point(data = top_stations_map, 
             aes(x = start_lng, y = start_lat, size = n, color = member_casual.x)) +
  theme_map() + labs(title = "Top 25 Start Stations by number of rides", size = "Number of rides", color = "Casual or Member")

```

## Share
I took the insights and visualizations I got from this analysis and put it together in a slide deick which you can see right [here](https://docs.google.com/presentation/d/1zroxVuxQpASXDeU32ZB6yL7w3FgSuwzUTALP_4CoLl0/edit?usp=sharing)
